---
title: "Untitled"
author: "Neville Andrade"
date: "9 April 2017"
output: 
  md_document:
    variant: markdown_github
---

## The goal


In this first chapter lets learn how to display and interpret data using some common display tools using R. Lets use the Mutual Funds.xls data which contains data on 259 mutual funds with some qualitative information on the fund type, objectives and risk buckets but also gives some quantitative performance metrics like 3 Year and 5 Year returns and best and worst quarter returns.

Lets start our journey into understanding data with the simplest plot for quantitative information which is the stem and leaf plot. 

## Presenting data

### Stem and Leaf plot

The stem and leaf plot is a simple plot which separates the data entries into leading digits or stems and trailing digits or leaves. The plot basically shows how the values distribute and cluster over the range of the observations in the data set.

In R there is a function called stem which is very suitable to display this data. It takes in 4 arguments out of which the first 2 are the most important. The first argument is an ordered vector of a variable and the second argument controls the height of the plot.

```{r}
library(xlsx)

mf_return <- read.xlsx(file = "./data/Mutual Funds.xls", sheetName = "Data")
head(mf_return)



# Order the 5 Year returns of the 259 Mutual funds
mf_order_5yr_return <- mf_return$X5.Yr.Return[order(mf_return$X5.Yr.Return)]

mf_order_5yr_return

stem(mf_order_5yr_return, scale= 2)

```

#### Interpretation

There are 4 mutual funds delivering -ve returns and 7 delivering returns > 20%. Most of the funds have returns between 8% to 12% returns. 

#### What is the stem and leaf graph good for

- It separates each data into groups: **Separation**
- It shows asymmetry in the data: **Assymetry**
- It visually shows irregularities: **Irregularities**
- It shows where most of the data is centered: **Centering**


#### What does the stem and leaf graph not tell you

1. Its not easy to visualize categorical plots. For example if one wants to visualize the 5 year returns based on mutual fund objectives

```{r}
head(mf_return)

#Get the levels of the Objectives: Growth and Value
levels(mf_return$Objective)


temp <- mf_return$X5.Yr.Return[mf_return$Objective %in% "Growth"]
(mf_return_5Yr_Growth <- temp[order(temp)])

temp <- mf_return$X5.Yr.Return[mf_return$Objective %in% "Value"]
(mf_return_5Yr_Value <- temp[order(temp)])

stem(mf_return_5Yr_Growth, scale=2)

stem(mf_return_5Yr_Value, scale=2)

```

You see that its not visually easy to compare the two plots.

2. The stem and leaf plots are fine for small data sets. Its better to use histograms for larger datasets.
3. For large data sets the center of the distribution is not clearly visible.


### Histogram

When there are a large number of observations, its useful to summarise the data using a histogram. A histogram is used to describe numerical data that have been grouped into frequency, relative frequency or percentage distributions. The variable of interest is plotted along the x axis grouped in intervals and the number of or percentage or proportion is plotted along the y axis.

Lets now construct a histogram of the 5 year returns of mutual funds data. 

```{r, label="Fig-Histogram"}
library(ggplot2)
ggplot(mf_return, aes(X5.Yr.Return)) + geom_histogram()

```

This plot shows the number of observations on the vertical y axis and the 5 yr return on the horizontal x axis. The plotting has been performed with a default bin width = 30.

```{r, label="Fig-Histogram-2"}
library(ggplot2)

ggplot(mf_return, aes(X5.Yr.Return)) + geom_histogram(bins = 50)

```

Manually modifying the number of bins from default of 30 to 50. I can also specify the binwidth or width of the class interval and enter it into the ggplot function. 

```{r}
length(mf_return$X5.Yr.Return)
range(mf_return$X5.Yr.Return)

# There are 259 observations in the range of -6 to 26. If I need say 100 bins

```

$$

Width\space of\space inteval=\frac{range}{number of desired class groupings} \\
 
 
Width\space of\space inteval=\frac{26-(-6)}{100} \\

Width\space of\space interval=\frac{32}{100}=0.32


$$

Plugging binwidth = 0.32 into the geom_histogram arguments gives the following

```{r, label="Fig-Histogram-3"}
ggplot(mf_return, aes(X5.Yr.Return)) + geom_histogram(binwidth = 0.32)
```

### Some of the additional things one can do with geom_histogram

Lets beautify the plot a bit. Lets make the colour of the bars blue and with a white border around them.

```{r, label="Fig-Histogram-4"}
ggplot(mf_return, aes(X5.Yr.Return)) + geom_histogram(binwidth = 0.32, fill="blue", col="white")
```


Lets make the bars colourful with a spectrum palette so that higher value show a different hue versus lower values

```{r, label="Fig-Histogram-5"}
ggplot(mf_return, aes(X5.Yr.Return)) + 
  geom_histogram(binwidth = 0.32, col="white", aes(fill=..count..))
```


Lets put another colour pallette

```{r, label="Fig-Histogram-6"}
ggplot(mf_return, aes(X5.Yr.Return)) + 
  geom_histogram(binwidth = 0.32, col="white", aes(fill=..count..)) +
  scale_fill_gradient("Count", low = "green", high = "red")

```

Lets add a title, labels to the axes and limit the scope of the graph
  
```{r, label="Fig-Histogram-7"}
ggplot(mf_return, aes(X5.Yr.Return)) + 
  geom_histogram(binwidth = 0.32, fill="blue", col="white")  + 
  labs(title="Histogram for Mutual fund returns") +
  labs(x="5 Yr return", y="Count of funds") + 
  xlim(c(-10,30)) +
  ylim(c(0,11))

```
  
Lets add a density plot over the graph

  
```{r, label="Fig-Histogram-8"}
ggplot(mf_return, aes(X5.Yr.Return)) + 
  geom_histogram(aes(y=..density..),binwidth = 0.32, fill="blue", col="white")  + 
  geom_density(col=2) +
  labs(title="Histogram for Mutual fund returns") +
  labs(x="5 Yr return", y="Count of funds") +
  xlim(c(-10,30)) 

```

Here I see that the frequency plot is done of the data 

### How do you observe the data being plotted by geom_histogram?

Use ggplot_build

```{r}
p <- ggplot(mf_return, aes(X5.Yr.Return)) + geom_histogram(binwidth = 0.32)

p_build <- ggplot_build(p)

p_build$data[[1]]

p_build$data[[1]]['count']


write.csv(p_build$data[[1]], file = "checkhistogram.csv")

```



### Barplot

What about a bar plot? The bar plot will plot each and every point of the data rather than binning the data in ranges.

```{r, label="Fig-Bar"}

ggplot(mf_return, aes(X5.Yr.Return)) + geom_bar()

```

This plot shows the count of each and every point on the vertical y axis versus the 5 yr return on the horizontal x axis. There is no grouping of data on the x axis which makes it seem very crowded.


### But what is the difference between a bar plot and a histogram?

A bar plot represents the data in the form of a categorical variable. Each column represents a categorical variable and the height of the column can either represent some quantitative measure or count of observations.  

On the other hand a histogram represents the data in the form of a quantitative variable. Each column represents a quantitative variable or a range of quantitative variables (class intervals) and the height of the column can either represent another quantitative measure or count of observations.  

So what I have plotted above in terms of the bar plot is not correct. I will later create one which makes sense. Right now lets try to make more sense of the histogram plot.

### How do you go about creating class intervals for the histogram

